---
title: Quick Control Charts for AFL
author: 'Robert Nguyen'
date: '2019-03-30'
slug: quick-control-charts-for-afl
categories: []
tags:
- fitzRoy
- betting
- graphics
description: 'Using control charts for AFL'
---

Who doesn't like a wikipedia entry [control chart](https://en.wikipedia.org/wiki/Control_chart) `If analysis of the control chart indicates that the process is currently under control (i.e., is stable, with variation only coming from sources common to the process), then no corrections or changes to process control parameters are needed or desired ` I mean gee whiz this sure could relate to something like I don't know AFL total game scores?

There seems to always be talk about the scores in AFLM see [AFL website](https://www.afl.com.au/news/2019-03-25/lowscoring-round-one-makes-unwanted-history), [foxsports](https://www.foxsports.com.au/afl/scorings-kickin-the-behind-proof-the-intention-behind-new-afl-rule-has-backfired-writes-brenton-speed/news-story/6aaf82e638044a8c42382d918cd5d3f1) just to name a couple. Of course you could find [more](https://myaccount.news.com.au/sites/cairnspost/subscribe.html?sourceCode=CPWEB_WRE170_a_GGN&mode=premium&dest=https://www.cairnspost.com.au/sport/afl/afl-daily-live-rolling-footy-news-from-around-australia-for-march-25-2019/live-coverage/7133f310052bc440fc77a24fec1ba888&memtype=anonymous) if you searched out for it as well. 

Let's use [fitzRoy](https://github.com/jimmyday12/fitzRoy) and the good people over at [statsinsder](https://www.statsinsider.com.au) who have kindly provided me with the expected score data you can get from the herald sun. 

First thing, lets use [fitzRoy](https://github.com/jimmyday12/fitzRoy)

```{r}
library(fitzRoy)
library(tidyverse)
library(ggQC)
fitzRoy::match_results%>%
  mutate(total=Home.Points+Away.Points)%>%
  group_by(Season,Round)%>%
  summarise(meantotal=mean(total))%>%
filter(Season>1989 &  Round=="R1")%>%
ggplot(aes(x=Season,y=meantotal))+geom_point()+geom_line()+stat_QC(method="XmR")+ylab("Mean Round 1 Total for Each Game") +ggtitle("Stop Freaking OUT over ONE ROUND")

```

So if we were to look at the control chart just for round 1 in each AFLM season since the 90s it would seem as though that even though this round was lower scoring that there isn't much too see here. 

After all we can and should expect natural variation in scores, wouldn't footy be boring if scores were the same every week. 

So next lets thing about the [expected scores framework](https://www.optasports.com/services/analytics/advanced-metrics/)

So the thing is we know that [week to week](http://www.matterofstats.com/mafl-stats-journal/2016/6/18/goal-kicking-accuracy-after-wins-and-losses) thanks to [matterofstats](https://twitter.com/MatterOfStats) we know that scoring does have certain properties. But what I want to look at with a control chart is last week in round 1 2019, were team creating worse oppotunities and outside the bounds that would cause us to worry if we were using a control chart at work. 



```{r echo=FALSE}
library(readxl)
CDExp_17 <- read_excel("~/Downloads/CDExp_17.xlsx")
head(CDExp_17)
```


So now lets get plotting.

```{r}
CDExp_17%>%
  mutate(totalexpected=H_CD_Exp + A_CD_Exp)%>%
  group_by(Year, Round)%>%
  summarise(meanexpected=mean(totalexpected))%>%
  unite("year_round", Year, Round)%>%
  ggplot(aes(x=year_round, y=meanexpected, group=1))+
  geom_line()+ 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

So what we can see here is that the plot has been re-ordered even though our dataframe is in the order that we want. 

```{r}
CDExp_17%>%
  mutate(totalexpected=H_CD_Exp + A_CD_Exp)%>%
  group_by(Year, Round)%>%
  summarise(meanexpected=mean(totalexpected))%>%
  unite("year_round", Year, Round)

```

So what is going on here is that ggplot orders the graph by the levels of our dataframe

```{r}
df<-CDExp_17%>%
  mutate(totalexpected=H_CD_Exp + A_CD_Exp)%>%
  group_by(Year, Round)%>%
  summarise(meanexpected=mean(totalexpected))%>%
  unite("year_round", Year, Round)
str(df)
df$year_round<-as.factor(df$year_round)
levels(df$year_round)
```

Which if you notice is the oder of our plot earlier. 

So how do we fix that? 

Lets just add a numeric column that is the row numbers and that will become our x-axis.

```{r}
CDExp_17%>%
  mutate(totalexpected=H_CD_Exp + A_CD_Exp)%>%
  group_by(Year, Round)%>%
  summarise(meanexpected=mean(totalexpected))%>%
  unite("year_round", Year, Round)%>%
  mutate(id = row_number())%>%
  ggplot(aes(x=id, y=meanexpected, group=1))+
  geom_line()+ 
  stat_QC(method="XmR")
  
```

There you go, if we think about expected scores as oppotunities created, round 1 was actually on the up from the finals series, players just didn't kick that well. But hey as we know goal kicking isn't going to happen at 100% of the time.
