---
title: Control Charts Another Package
author: Robert Nguyen
date: '2019-04-20'
slug: control-charts-another-package
categories:
  - viz
tags:
  - viz
description: 'xmrr'
---

I got an email from Alex Zanidean, who runs the [xmrr](https://cran.r-project.org/web/packages/xmrr/xmrr.pdf) package 

"You might enjoy my package xmrr for similar charts – but mine recalculate the bounds automatically" and if we go to the vingette, "XMRs combine X-Bar control charts and Moving Range control charts. These functions also will recalculate the reference lines when significant change has occurred" This seems like a pretty handy thing. So lets do it.

First lets do our graphic from our previous post using [ggQC](https://cran.r-project.org/package=ggQC)

```{r}
library(fitzRoy)
library(tidyverse)
library(ggQC)
library(xmrr)
fitzRoy::match_results%>%
  mutate(total=Home.Points+Away.Points)%>%
  group_by(Season,Round)%>%
  summarise(meantotal=mean(total))%>%
filter(Season>1989 &  Round=="R1")%>%
  ggplot(aes(x=Season,y=meantotal))+geom_point()+geom_line()+stat_QC(method="XmR")+ylab("Mean Round 1 Total for Each Game") +ggtitle("Stop Freaking OUT over ONE ROUND")

df<-fitzRoy::match_results%>%
  mutate(total=Home.Points+Away.Points)%>%
  group_by(Season,Round)%>%
  summarise(meantotal=mean(total))%>%
filter(Season>1989 &  Round=="R1")

```

So when using a package for the first time, one of the best things about the R community is how the examples are usually fully reproducible and this helps.

From the [github](https://github.com/Zanidean/xmrr)

```{r}
Year <- seq(2001, 2009, 1)
Measure <-  runif(length(Year))

df <- data.frame(Year, Measure)
head(df)

xmr(df, "Measure", recalc = T)
```

Lets create a similar dataframe as df, but using data from [fitzRoy](https://github.com/jimmyday12/fitzRoy)

```{r}
df<-fitzRoy::match_results%>%
  mutate(total=Home.Points+Away.Points)%>%
  group_by(Season,Round)%>%
  summarise(meantotal=mean(total))%>%
filter(Season>1989 &  Round=="R1")%>%
  select(Season, meantotal)
df<-data.frame(df)
xmr_data <-xmr(df, "meantotal", recalc = T)

xmr_chart(df = xmr_data, 
          time = "Season", 
          measure = "meantotal",
          line_width = 0.75, text_size = 12, point_size = 2.5)

```

Does this tell a different story or a very similar one to earlier?
