---
title : "Joining Betting Data"
description : "Joining on Betting Data to fitzRoy"
author : "Robert Nguyen"
date : 2018-06-27
tags : ["betting","fitzRoy"]

---


This example will be using [Betfair](https://www.betfair.com.au/hub/tools/betting-tools/betfair-data/) which we hope to add to fitzRoy in the future. Making it easier to compare [Squiggle](http://squiggle.com.au) and maybe your own models vs the market!

# Step One Read in the Betfair Data

The first step is just go to the [betfair site](https://www.betfair.com.au/hub/tools/betting-tools/betfair-data/) and download the AFL file with the data we need.

I will only be joining the 2017-onwards data because that's what squiggle has!

```{r}
library(readxl)
library(tidyverse)



filetodownload <- "https://www.betfair.com.au/hub/wp-content/uploads/sites/2/2018/10/Copy-of-Weekly-AFL-Data-Dump-2017-YTD-2.xlsx"
download.file(filetodownload,"aflodds.xlsx",mode="wb")

Copy_of_Weekly_AFL_Data_Dump_2017_YTD <- read_excel("aflodds.xlsx", 
    col_types = c("date", "text", "numeric", 
         "text", "text", "text", "text", "numeric", 
        "text", "text", "text", "text", "text", 
         "text", "text", "text"))
dim(Copy_of_Weekly_AFL_Data_Dump_2017_YTD)
head(Copy_of_Weekly_AFL_Data_Dump_2017_YTD)

df_bookie<-slice(Copy_of_Weekly_AFL_Data_Dump_2017_YTD,2:1430)
names(df_bookie) <- lapply(df_bookie[1, ], as.character)
names(df_bookie)

```

The columns of interest to join on would be, the Date, the selection name and WAP and we want to maybe join this on to the  [squiggle guys](http://squiggle.com.au). The next thing we notice by viewing the dataset `View(Copy_of_Weekly_AFL_Data_Dump_2017_YTD)` is that if we look at the column `INPLAY` we want when it is `N` the reason being is that we want to compare the squiggle pre game tips with what the market was doing pre game.

# Step Two Get the squiggle data

```{r}
library(fitzRoy)
tips <- get_squiggle_data("tips")
df<-tips%>%mutate(home.margin=ifelse(hteam==tip, margin,-margin))%>%
  mutate(away.margin=ifelse(ateam==tip, margin,-margin)) %>%
  select(source,date,correct,  hconfidence,hteam,
         ateam,home.margin,away.margin,err ,tip,round, year)
 

df1<-select(df,source, date, correct, hconfidence,hteam, home.margin, err, tip, round, year )
df1$H_A<-"HOME"
df2<-select(df, source, date, correct, hconfidence, ateam, away.margin, err, tip, round, year)
df2$H_A<-"AWAY"
colnames(df1)[5]<-"Team"

colnames(df1)[6] <- "margin"

colnames(df2)[5]<-"Team"

colnames(df2)[6]<-"margin"

df3<-rbind(df1,df2)
colnames(df3)[2]<-"Date"



```
# Step Three make sure the names of columns align!


```{r}
colnames(df_bookie)[1] <- "Date"
colnames(df_bookie)[10] <- "Team"
colnames(df_bookie)[3]<-"x3"
colnames(df_bookie)[8]<-"x8"
df_bookie<-df_bookie%>%filter(INPLAY=="N")%>%
  select(Date, Team, WAP)
df_bookie$Date<-as.Date(df_bookie$Date)
df3$Date<-as.Date(df3$Date)

```

# Step 4 Check what you are joining by!

Before we join on the datasets, we need to make sure that our IDs line up, so that our teams are named the same in *both* datasets. 



```{r}
unique(df_bookie$Team)
unique(df3$Team)

```


From here we can see that not all the team names are aligned so we need to fix them up, as an example we have the Adelaide Crows in 2017 and Adelaide in 2018. When we try and join to Squiggle we will miss out some games in 2017 because its trying to match Adelaide (from squiggle) with Adelaide Crows from betfair. 

To confirm this we can use below

```{r}

df_bookie%>%filter(Team %in% c("Adelaide Crows", "Adelaide"))



inner_join(df_bookie,df3, by=c("Team","Date"))%>%
  filter(Team %in% c("Adelaide Crows", "Adelaide"))
```


So lets just fix up the team names



```{r}


df_bookie<-df_bookie %>%
     mutate(Team=replace(Team,Team=="Adelaide Crows", "Adelaide")) %>%
   mutate(Team=replace(Team,Team=="Sydney Swans", "Sydney")) %>%
   mutate(Team=replace(Team,Team=="Gold Coast Suns", "Gold Coast")) %>%
   mutate(Team=replace(Team,Team=="GWS Giants", "Greater Western Sydney")) %>%
   mutate(Team=replace(Team,Team=="GWS", "Greater Western Sydney")) %>%
   mutate(Team=replace(Team,Team=="West Coast Eagles", "West Coast")) %>%
   mutate(Team=replace(Team,Team=="Brisbane", "Brisbane Lions")) %>%
   mutate(Team=replace(Team,Team=="Geelong Cats", "Geelong")) %>%
   mutate(Team=replace(Team,Team=="Melbourne Demons", "Melbourne")) %>%
       as.data.frame()


df_joined<-left_join(df3,df_bookie, by=c("Team","Date"))

```

# Step 5 Check the Join

So remember what we wanted originally, we wanted to join betting data to the squiggle data. Our squiggle data was a dataframe called `df3` and our bookie data was a dataframe called `df_bookie`. One way to check our join, is to make sure the dimension lines up. 

We do this using `dim`.



```{r}
dim(df3)

dim(df_bookie)
dim(df_joined)

```

Our dimension for df3 is 7842 rows and 11 columns, and our joined data `df_joined` is 7842 rows and 12 columns, i.e. it is the same as df3 (squiggle data) but has an extra column (betfair odds)


So it seems as though we have a clean dataset to analyse. Now we just got to go ahead and do it!


# A quick analysis

Someone asked me who do you think tips underdogs the best? 

Well if you just wanted to know who has most winning underdog 'bets' (I don't know if any of the squigglers bet).


```{r}
df_joined%>%filter(WAP>2 &correct ==1 & Team==tip)%>%
  group_by(source)%>%
  summarise(count=n())%>%
  arrange(desc(count))

```


That tells me Footy Maths institute had the most winning underdog bets, but just how many times did they have to bet?


```{r}
df_joined%>%filter(WAP>2 & Team==tip)%>%
    group_by(source, correct)%>%
    summarise(count=n())%>%
    group_by(source)%>%
    mutate(total_bets=sum(count), pert=count/total_bets)%>%
    filter(correct==1)%>%
    arrange(desc(pert))

```

Note we can automatically download the AFL file using the following script.

```{r, eval=FALSE}
library(readxl)
filetodownload <- "http://www.aussportsbetting.com/historical_data/afl.xlsx"
download.file(filetodownload,"aflodds.xlsx",mode="wb")


res <-read_excel("aflodds.xlsx", sheet = 1,skip = 1) 



```



