---
title : "Data Viz and Manipulation P2"
description : "A newbies guide to AFL visualisation with R P2"
author : "Robert Nguyen"
date : 2018-05-20
tags : ["fitzRoy", "graphics","R","open science" , "tidyverse"]

---

# Using fitzRoy and the tidyverse

```{r}
library(fitzRoy)
library(tidyverse)
```

So what we did before, was we created 3 variables from afltables about Tony Lockett. They were Year (afl season), GM (totals games played in afl season) and GL (total goals kicked in season).  

These are summary statistics i.e. the year summaries (totals) of games played and goals kicked. 

Something you might have noticed with `fitzRoy` is that the afldata is stored as game by game data and not season summaries. 

So how do we create these summary statistics? 

# Step One - Get the dataframe with it all

Data is stored in different places in fitzRoy, afldata contains the data we are after at the moment

`head(df)` 

```{r}
options(max.print=100)
df<-fitzRoy::afldata


```

# Step Two - Select the columns that you want [`select`](https://dplyr.tidyverse.org/reference/select.html)

```{r}
df%>%
  select(First.name, Surname, Season, Goals)

```

# Step Three - Create the grouped data [`group_by`](https://dplyr.tidyverse.org/reference/group_by.html)

Now this might seem a bit weird, because on first look it would seem as though group_by doesn't do anything. 

```{r}

df%>%
  select(First.name, Surname, Season, Goals)%>%
  group_by(Season)
```

It doesn't look like our dataset has changed, but if we look closely we have a new `Groups:   Season [121]` and hopefully what you have noticed is that our dataset `afldata` has 121 unique values. Which we can check below. 

```{r}
length(unique(df$Season))
```

What this does is adds a grouping structure to our data, which means instead of operations acting element wise like they did earlier now operations can happen by group. 

So why would you want to group data? Well lots of interesting things are done on a `group_by` basis. For example we might be interested in goal kicking trends by [`Season`](https://www.foxsports.com.au/afl/afl-goalkicking-continues-to-be-under-scrutiny-as-league-accuracy-falls-to-its-worst-since-1990/news-story/b01d958a5952a86e714e7ef0a41f75d5)
Or 
* by [`round`](https://twitter.com/MatterOfStats/status/997814414832095237) 
* by [`Playing.For`](https://twitter.com/MatterOfStats/status/997814414832095237)
* by [opponent](https://twitter.com/MatterOfStats/status/998017464519675904)
* by [wins and losses](https://twitter.com/MatterOfStats/status/998018492426141696)

We can also `group_by` more than one thing. For example what if we wanted by `Season` and `Round`

```{r}
df%>%
  select(Season,Round, Goals, Behinds)%>%
  group_by(Season, Round)
```
What we can now see is that we have grouped by 2 variables (Season, Round) and have formed 769 groups. `Groups:   Season, Round [2,769]`

# Step Four - Take the grouped data and `summarise` it

So recall earlier we grouped our data by year so lets summarise it. 

```{r}
df%>% 
  select(First.name, Surname, Season, Goals)%>%
  group_by(Season)%>%
  summarise(total_goals=sum(Goals))
```
Ok that doesn't quite seem like what we were thinking we wanted the `total_goals` of Tony Lockett by season, but all we got was Season total goals. 

Well duh! We only did a `group_by` by `Season`, we should have done it by `Season, First.name, Surname` should we have? 

Lets check.

```{r}
df%>%
  select(Season, First.name, Surname, Goals)%>%
  group_by(Season, First.name, Surname)%>%
  summarise(total_goals=sum(Goals))
  
```
This looks a lot better, only issue is that we don't see Tony Lockett but we see all these other blokes like Alb Thomas, he didn't kick 1,360 snags. 

# Step five - `filter` the data

```{r}
df%>%
  select(Season, First.name, Surname, Goals)%>%
  group_by(Season, First.name, Surname)%>%
  summarise(total_goals=sum(Goals))%>%
  filter(First.name=="Tony" , Surname=="Lockett")
```

Now I think I might know what you are thinking. "Weren't we supposed to be plotting average goals by year?"

To do that all we have to do is change `sum` to `mean` and `total_goals` to `average_goals` (these names can be anything but its helpful if they are descriptive).

```{r}
df%>%
  select(Season, First.name, Surname, Goals)%>%
  group_by(Season, First.name, Surname)%>%
  summarise(average_goals=mean(Goals))%>%
  filter(First.name=="Tony" , Surname=="Lockett")
```

# Step Six `ggplot` our way to a glory

```{r}
df%>%
  select(Season, First.name, Surname, Goals)%>%
  group_by(Season, First.name, Surname)%>%
  summarise(average_goals=mean(Goals))%>%
  filter(First.name=="Tony" , Surname=="Lockett")%>%
  ggplot(aes(x=Season, y=average_goals)) +geom_line(color="red") +ggtitle("Tony Lockets Average Goals Per Game by Age")

```

