---
title : "A Different Free Kick Ladder"
description : "Doing a different ladder to Triple M"
author : "Robert Nguyen"
date : 2018-06-25
tags : ["fitzRoy"]

---

Each week [triplemfooty](https://twitter.com/triplemfooty/status/1011093422608867329) tweets out a free kick ladder that is based only on total free kick differential throughout the season. 

But what if we did a ladder in the same style of the home and away ladder. 

* 4 points for a win (winning free kick count in game)
* 2 points for a draw (drawn free kick count in game)
* 0 points for a loss (losing the free kick count in game)
* percentage which is $100 * (totalfreesfor/totalfreesagainst)$

Well thankfully to the new afl R package [fitzRoy](https://github.com/jimmyday12/fitzRoy) that is possible!


```{r}
library(tidyverse)
fitzRoy::get_footywire_stats(9514:9630)%>%
    group_by(Round, Team)%>%
  mutate(Free_diff=sum(FF)-sum(FA), 
  sumFF=sum(FF), sumFA=sum(FA))%>%
  select(Team, sumFF, sumFA, Free_diff)%>%
  distinct()%>%
  mutate(points=if_else(Free_diff>0,4,if_else(Free_diff==0,2,0)))%>%
  group_by(Team)%>%
  summarise(total_points=sum(points),
  percentage_Frees=100*(sum(sumFF)/sum(sumFA)))%>%
  arrange(desc(total_points), desc(percentage_Frees))

```

So there has been a lot of [talk](https://www.sen.com.au/news/2018/06/05/afl-cant-ignore-eagles-free-kicks-says-former-umpires-boss/) [recently](https://www.foxsports.com.au/afl/why-home-teams-get-more-free-kicks-unbalanced-free-kick-counts-explained/news-story/a212665ad4d54849f18119f3d88ebca3) about how [West Coast](https://www.perthnow.com.au/sport/west-coast-eagles/adam-simpson-has-no-answer-for-west-coast-eagles-frees-advantage-at-optus-stadium-ng-b88881644z) receive a free kick advantage at home. The media hype has got to the point where the [AFL will investigate](https://www.heraldsun.com.au/sport/afl/teams/west-coast/west-coasts-free-kick-domination-puzzles-afl-umpires/news-story/febbc87b37b79a945d890ddc0898629b).

So how can we use this ladder concept to see if there might be something there?

Lets first start by looking at the free kick ladder for 2018.

```{r}
library(tidyverse)
fitzRoy::get_footywire_stats(9514:9630)%>%
    group_by(Round, Team)%>%
  mutate(Free_diff=sum(FF)-sum(FA), 
  sumFF=sum(FF), sumFA=sum(FA))%>%
  select(Team, sumFF, sumFA, Free_diff)%>%
  distinct()%>%
  mutate(points=if_else(Free_diff>0,4,if_else(Free_diff==0,2,0)))%>%
  group_by(Team)%>%
  summarise(total_points=sum(points),
  percentage_Frees=100*(sum(sumFF)/sum(sumFA)))%>%
  arrange(desc(total_points), desc(percentage_Frees))

```

So West Coast do well on frees but certainly they are not top of the comp. 

But the perception is that West Coast are being favoured at home, so lets divide up this table into home and away.

```{r}
df<-fitzRoy::get_footywire_stats(9514:9630)%>%
  group_by(Round, Team, Status)%>%
  mutate(Free_diff=sum(FF)-sum(FA), 
         sumFF=sum(FF), sumFA=sum(FA))%>%
  select(Team, sumFF, sumFA, Free_diff)%>%
  distinct()%>%
  mutate(points=if_else(Free_diff>0,4,if_else(Free_diff==0,2,0)))%>%
  group_by(Team,Status)%>%
  summarise(total_points=sum(points),
            percentage_Frees=100*(sum(sumFF)/sum(sumFA)))%>%
  arrange(desc(total_points), desc(percentage_Frees))

df%>%
  filter(Status=="Home")%>%
   arrange(desc(total_points), desc(percentage_Frees))

df%>%
  filter(Status=="Away")%>%
   arrange(desc(total_points), desc(percentage_Frees))


```

Looking at our tables for Home vs Away, it would seem as though for this year at least that Adelaide is being favoured more, they lead the home free kick table are 4th on the overall table and just outside the eight on the away table. What's the go with North Melbourne they lead the overall table and the away table but finish a mere 7th for games at home. 

What about those tigers? For all the talk about home ground advantage and there being some influence on umpires it doesn't seem to bother them if its 80,000 screaming tigers fans. Could the noise of affirmation over in the west, be louder than the tiger roar?

Other people might want to see a plot of free kicks differential but graphically. So lets do that. 

```{r}
fitzRoy::player_stats%>%
  select(-Player)%>%
  mutate(Round = parse_number(Round))%>%
  mutate(Season_round=Round+Season)%>%
  group_by(Season_round,Season, Round, Venue, Team, Status, Match_id,Date)%>%
  summarise(Fdiff=sum(FF)-sum(FA))%>%
  ggplot(aes(x=Season_round, y=Fdiff))+
  geom_point(aes(colour = factor(Status))) + 
  geom_hline(yintercept = 0)+
  facet_wrap(~Team)

```


Visually if we look at the WCE plot it does look as though they get the rub of the green at home but not away. 

Are they just making less mistakes at home?


```{r}
fitzRoy::player_stats%>%
  select(-Player)%>%
  mutate(Round = parse_number(Round))%>%
  mutate(Season_round=Round+Season)%>%
  group_by(Season_round,Date,Season, Round, Team, Status, Opposition, Venue, Match_id)%>%
  summarise_all(.funs = sum)%>%
  group_by(Match_id)%>%
  arrange(Match_id)%>%
  mutate_if(is.numeric, funs(difference=c(-diff(.), diff(.))))%>%
  ggplot(aes(x=Season_round,y=CG_difference))+
  geom_point(aes(colour = factor(Status))) + geom_hline(yintercept = 0)+
  facet_wrap(~Team)

```

```{r}
fitzRoy::player_stats%>%
  select(-Player)%>%
  mutate(Round = parse_number(Round))%>%
  mutate(Season_round=Round+Season)%>%
  group_by(Season_round,Date,Season, Round, Team, Status, Opposition, Venue, Match_id)%>%
  summarise_all(.funs = sum)%>%
  group_by(Match_id)%>%
  arrange(Match_id)%>%
  mutate_if(is.numeric, funs(difference=c(-diff(.), diff(.))))%>%
  ggplot(aes(x=Season_round,y=CG_difference))+
  geom_point(aes(colour = factor(Status))) + geom_hline(yintercept = 0)+
  facet_wrap(~Team)

```

Again visually it looks like the eagles are making less mistakes at home. 

What about their tackle differential?



```{r}

fitzRoy::player_stats%>%
  select(-Player)%>%
  mutate(Round = parse_number(Round))%>%
  mutate(Season_round=Round+Season)%>%
  group_by(Season_round,Date,Season, Round, Team, Status, Opposition, Venue, Match_id)%>%
  summarise_all(.funs = sum)%>%
  group_by(Match_id)%>%
  arrange(Match_id)%>%
  mutate_if(is.numeric, funs(difference=c(-diff(.), diff(.))))%>%
  ggplot(aes(x=Season_round,y=T_difference))+
  geom_point(aes(colour = factor(Status))) + geom_hline(yintercept = 0)+
  facet_wrap(~Team)

```

Doesn't look like their is much difference, so for WCE it looks as though they do get more free kicks at home vs away, they make less mistakes at home but tackles don't seem to change by home/away status. 


So while these plots paint a picture, we should always dig in a bit more to see if there is something there, or if it is just the mind trying to spot a pattern when there is none. 

