---
title : "But We Won Everywhere but the Scoreboard"
description : "Introduction to logistic regression"
author : "Robert Nguyen"
date : 2018-06-23
tags : ["regression", "fitzRoy","modelling"]

---

Something that gets to many a footy fan, is the feeling that your team has won the game in most areas expect on the scoreboard. 

Thinking about this statement a little bit deeper has the following implication. That there are some areas of the game, that if you win, you tend to win the game. 

We can think about this by putting it into a model. 


The data we will use here is the game differentials of the various game statistics that are available using [fitzRoy](https://github.com/jimmyday12/fitzRoy)


The general idea is its fairly obvious if you score more than the opposition you win. Shocking I know! What we are trying to do here, is come up with a concept of we won other things but not the scoreboard, usually when we win these things we tend to win a game. 

So we know thanks to [fitzRoy](https://github.com/jimmyday12/fitzRoy) we have access to the data on [footywire](https://www.footywire.com/), which has a few extra variables that are not on [afltables](https://afltables.com/afl/afl_index.html). 

So imagine this scenario, you don't have access to the scores, only the in game statistics as per [these](https://www.footywire.com/afl/footy/ft_match_statistics?mid=9626) [pages](https://www.footywire.com/afl/footy/ft_match_statistics?mid=9626&advv=Y), what variables/differentials would you look at to decide to will win? 

We know that for example, looking only at [free](https://analysisofafl.netlify.com/makemeauser/2018-06-25-a-different-free-kick-ladder/) [kick](https://www.triplem.com.au/sport/afl/news/afl-free-kick-differential-ladder-round-14-2018) differential isn't very predictive. Is there a combination of differentials that when a team wins they are more likely to win? Lets build a model to find out. 


The model I will use here, is a logistic regression model with the binary outcome being win/loss. 

# Step One Build the dataset

```{r}
library(tidyverse)
fitzRoy::player_stats%>%
    filter(Season>2014)%>%
    select(-Player)%>%
    group_by(Date,Season, Round, Team, Status, Opposition, Venue, Match_id)%>%
    summarise_all(.funs = sum)%>%
group_by(Match_id)%>%
  arrange(Match_id)%>%
  mutate(diff_cp=c(-diff(CP), diff(CP)))

```


Something to remember with fitzRoy is that footywire doesn't have tackles inside 50, meters gained etc for games previous to 2015, so this becomes our first `filter` 

The next thing we do, is we `select` all the columns except for Player, the reason being is that we don't need it and also its nice to see how to "deselect" columns as well as select them. 

Remember we want data at a game level for each team and we want to be able to come up with the in game differential, this is where our next `group_by` comes in handy and its related to the `summarise_all`. 

You might be thinking jeez mate that's a lot to group by can't you just use Match_id as they are already unique. Yes that is true, I could have, but one thing about `summarise_all` is that is summarises every column that is not in the `group_by`. You can test this out by running below, it should spit out an error message. `Error in summarise_impl(.data, dots) : 
  Evaluation error: invalid 'type' (character) of argument.`
  
```{r, eval=FALSE}

fitzRoy::player_stats%>%
    filter(Season>2014)%>%
    select(-Player)%>%
    group_by( Match_id)%>%
    summarise_all(.funs = sum)
```

After the `summarise_all`, we group by `Match_id` so we can find the differentials by `Match_id` so we can come up with as an example the contested possession differential `diff_cp`. I like to `arrange` the dataset so I can do sanity checks vs [footywire](https://www.footywire.com/afl/footy/ft_match_statistics?mid=9626).

So the next thing you might be thinking, is jeez doing that for all variables finding their differentials seems a bit tedious. 

That is where we can use `mutate_if`, we can just `mutate` all columns that `is.numeric` and is not in the `group_by` which is another way we could have done original `summarise_all` instead we could have used `summarise_if` BUT match_id is numeric!

You can check this by using 
```{r}
str(fitzRoy::player_stats)
```

So hopefully you have seen a couple of short cuts that come from knowing the dataset and this is an example of why its important to do your checks! 

Here is a great a [guide to bad data](https://github.com/Quartz/bad-data-guide) which provides a nice series of steps in how to check things that might commonly go wrong. 




```{r}
fitzRoy::player_stats%>%
         filter(Season>2014)%>%
         select(-Player,-Date)%>%
         group_by(Season, Round, Team, Status, Opposition, Venue, Match_id)%>%
         summarise_all(.funs = sum)%>%
         group_by(Match_id)%>%
         arrange(Match_id)%>%
         mutate_if(is.numeric, funs(difference=c(-diff(.), diff(.))))

```

# Step Two - Joining Datasets

Now we could just sum up the goals by each player and their behinds, but this would miss out on rushed behinds. So lets join our [footywire](https://www.footywire.com) dataset with the [afltables](https://afltables.com/afl/seas/2018.html) match results



```{r}
df<-fitzRoy::player_stats%>%
         filter(Season>2014)%>%
         select(-Player)%>%
         group_by(Date,Season, Round, Team, Status, Opposition, Venue, Match_id)%>%
         summarise_all(.funs = sum)%>%
         group_by(Match_id)%>%
         arrange(Match_id)%>%
         mutate_if(is.numeric, funs(difference=c(-diff(.), diff(.))))

df2<-fitzRoy::match_results
df2<-df2%>%filter(Season>2014)
df3<-select(df2, Date, Round, Home.Team, Home.Points)
df4<-select(df2, Date, Round, Away.Team, Away.Points)
colnames(df3)[3]<-"Team"
colnames(df3)[4]<-"Points"
colnames(df4)[3]<-"Team"
colnames(df4)[4]<-"Points"


df5<-rbind(df4,df3)



df5<-df5 %>%mutate(Team = str_replace(Team, "Brisbane Lions", "Brisbane"))

df5<-df5 %>%mutate(Team = str_replace(Team, "Footscray", "Western Bulldogs"))


df6<-inner_join(df,df5, by=c("Team","Date"))

dataset_columns <- c(1,2,4,6,7,8,44:80,81)
dataset<-df6%>%group_by(Match_id)%>%
  arrange(Match_id)%>%
  mutate(Margin=c(-diff(Points), diff(Points)))%>%
  mutate(Win_loss=if_else(Margin>0,1,0,NULL))%>%
  select(dataset_columns)


```

# Taking Submissions. 

I asked twitter, [what areas do you think](https://twitter.com/anoafl/status/1011156299323604992) are important to win, to win a game. I got [a response](https://twitter.com/michael_timbs/status/1011159239232876544) and here we go!

So from our above script, all we need to be able to add in is goal accuracy as a predictor or goal accuracy differential? 



```{r}
df<-fitzRoy::player_stats%>%
         filter(Season>2014)%>%
         select(-Player)%>%
         group_by(Date,Season, Round, Team, Status, Opposition, Venue, Match_id)%>%
         summarise_all(.funs = sum)%>%
         group_by(Match_id)%>%
         arrange(Match_id)%>%
         mutate_if(is.numeric, funs(difference=c(-diff(.), diff(.))))

df2<-fitzRoy::match_results
df2<-df2%>%filter(Season>2014)
df3<-select(df2, Date, Round, Home.Team, Home.Points,Home.Goals,Home.Behinds)
df4<-select(df2, Date, Round, Away.Team, Away.Points,Away.Goals,Away.Behinds)
df3$Accuracy<-(df3$Home.Goals/(df3$Home.Goals+df3$Home.Behinds))
colnames(df3)[3]<-"Team"
colnames(df3)[4]<-"Points"
colnames(df3)[5]<-"Goals"
colnames(df3)[6]<-"Behinds"
df4$Accuracy<-(df4$Away.Goals/(df4$Away.Goals+df4$Away.Behinds))
colnames(df4)[3]<-"Team"
colnames(df4)[4]<-"Points"
colnames(df4)[5]<-"Goals"
colnames(df4)[6]<-"Behinds"


df5<-rbind(df4,df3)



df5<-df5 %>%mutate(Team = str_replace(Team, "Brisbane Lions", "Brisbane"))

df5<-df5 %>%mutate(Team = str_replace(Team, "Footscray", "Western Bulldogs"))

df6<-inner_join(df,df5, by=c("Team","Date"))


dataset_columns <- c(1,2,4,6,7,8,45:77,82:84)
dataset<-df6%>%group_by(Match_id)%>%
  arrange(Match_id)%>%
  mutate(Margin=c(-diff(Points), diff(Points)))%>%
  mutate(Win_loss=if_else(Margin>0,1,0,NULL))%>%
  select(dataset_columns)

```


# Building the Logistic Regression Model


```{r}
library(aod)
library(ordinal)
library(lme4)
in.sample  <- subset(dataset, Season %in% c(2015:2017))
#in.sample  <- subset(mydata, year ==2008)
out.sample <- subset(dataset, Season == 2018)

in.sample$Win_loss <- factor(in.sample$Win_loss)

out.sample$Win_loss<-factor(out.sample$Win_loss)


```

To know which columsn we want to scale an easy way is to go names(dataset) which should print out the column names.

```{r}

temp1<-scale(in.sample[,7:40])
in.sample[,7:40]<-temp1
#attributes(temp1)
temp1.center<-attr(temp1,"scaled:center")
temp1.scale<-attr(temp1,"scaled:scale")




```


```{r}


m <- glm(Win_loss ~I50_difference+
           Accuracy+
           R50_difference+
           CCL_difference+
           SCL_difference+
           MI5_difference , data = in.sample, family =binomial)
          
summary(m)


newdata   <- out.sample[ , -ncol(out.sample)]


newdata[,7:40]<-scale(newdata[,7:40],center=temp1.center,scale=temp1.scale) 



pre.dict    <- predict(m,newdata=newdata, type="response")
pre.dict.m  <- data.frame(matrix(unlist(pre.dict), nrow= nrow(newdata)))
colnames(pre.dict.m) <- c("prob.win")


newdata.pred  <- cbind.data.frame(newdata, pre.dict.m)
newdata.pred%>%
  select(Team, Opposition, Venue,  Margin, prob.win)%>%
  filter(Margin<0)%>%
  arrange(desc(prob.win))%>%
  top_n(10)

```

The good thing about having a template sorted out, is that you can make quick changes as you think of other variables you want to test. 

For example [JT](https://twitter.com/Trooa/status/1011406234207916032) asked about kicks, inside 50s, marks in 50, tackles in 50 and meters gained. 

Well lets look at his list of games

```{r}
m <- glm(Win_loss ~I50_difference+
           K_difference+
          I50_difference+
           MI5_difference+
           T5_difference+
           MG_difference, data = in.sample, family =binomial)
          
summary(m)


newdata   <- out.sample[ , -ncol(out.sample)]


newdata[,7:40]<-scale(newdata[,7:40],center=temp1.center,scale=temp1.scale) 



pre.dict    <- predict(m,newdata=newdata, type="response")
pre.dict.m  <- data.frame(matrix(unlist(pre.dict), nrow= nrow(newdata)))
colnames(pre.dict.m) <- c("prob.win")


newdata.pred  <- cbind.data.frame(newdata, pre.dict.m)
newdata.pred%>%
  select(Team, Opposition, Venue,  Margin, prob.win)%>%
  filter(Margin<0)%>%
  arrange(desc(prob.win))%>%
  top_n(10)
```




