---
title : "Regression to the Mean"
description : "An example using goalkicking"
author  : "Robert Nguyen"
date : 2018-06-20
tags : ["regression", "fitzRoy","modelling"]

---

We also see this in sports, Hollinger refers to it as the [fluke rule](http://www.espn.com/nba/columns/story?columnist=hollinger_john&page=All-Decline-Team-081031). Bill James calls it the [plexiglas principle](http://grantland.com/the-triangle/fantasy-baseball-nerdy-keys-to-victory-the-plexiglas-principle/), you might call it the [Sports Illustrated Jinx](https://en.wikipedia.org/wiki/Sports_Illustrated_cover_jinx). They all are different names from different people who have observed the same thing mean reversion or regression to the mean. 

Regression to the mean as a concept was first introduced in 1886 by Galton who observed that on average taller parents have tall kids but not as tall as them. In otherwords their heights 'shrunk' towards the overall average at the time. 

Regression to the mean is a powerful concept in statistics and how can we use it? 

The following will be a simple goal-kicking example but if you wanted to, you could use something like supercoach scores to beat your friends or the bookie. 


# Step 1 create the dataset

Thankfully we can use [fitzRoy](https://github.com/jimmyday12/fitzRoy) for the data we need pretty easily. Here we will take goals (G), behinds (B) and create a variable using `mutate` called `shots` which just represent total shots on goals (we don't have access to kicks that were out on the full or dropped short) so this becomes the sum of goals and behinds.  

We are looking at regression to the mean for goal kicking accuracy on a year to year bases on a player level. Hence our `group_by(Player, Team, Season)`

```{r}

library(tidyverse)
library(fitzRoy)
data<-fitzRoy::player_stats

data<-data%>%
  group_by(Player, Team, Season)%>%
  summarise(TG=sum(G), TB=sum(B))




data%>%
  group_by(Player,Team, Season)%>%
  mutate(shots= TG + TB)%>%
  ggplot(aes(x=shots))+geom_histogram(binwidth = 5)

data<-data%>%
  group_by(Player,Team, Season)%>%
  mutate(shots= TG + TB)
```

# Step Two Subsetting our data, finding ours means

Looking at our earlier plot we see that we have lots of players who have failed to take a single shot at goal. We don't want this to effect what we are doing down the track so lets arbitrary cut off shots at say 20 or more

```{r}
subset = data$shots > 19


good = data$TG[subset]
had_a_go = data$shots[subset]
names(good) = names(had_a_go) = paste(data$Player[subset], data$Season[subset])


S = good/had_a_go

muT = sum(good)/sum(had_a_go)
print(muT)

sigma2L = muT*(1-muT)/had_a_go

sigma2T = 0

for (i in 1:100000) {
  weight = 1/(2*(sigma2T + sigma2L)^2)
  sigma2T = sum(weight*((S - muT)^2 - sigma2L))/sum(weight)
}
print(sigma2T)


sqrt(sigma2T)


skill = (muT/sigma2T + S/sigma2L)/(1/sigma2T + 1/sigma2L)


head(sort(skill, decreasing = TRUE))
tail(sort(skill, decreasing = TRUE))

plot(S, skill, xlim = c(.2, .5), col = 'orange')
abline(0, 1, col = 'blue')


plot(density(S), ylim = c(0, 30), col = 'orange',
     main = 'Distribution of Goal kicking skill in AFL',
     xlab = 'Goal Kicking  percentage')
lines(density(skill), col = 'blue')
legend('topright', c('True skill', 'Observed'),
       col = c('blue', 'orange'), lty = 1)

```