---
title : "Data Viz and Manipulation P2"
description : "A newbies guide to AFL visualisation with R P2"
author : "Robert Nguyen"
date : 2018-05-21
tags : ["fitzRoy", "graphics","R","open science" , "tidyverse"]

---

# Using fitzRoy and the tidyverse

```{r}
library(fitzRoy)
library(tidyverse)
```

So what we did before, was we created 3 variables from afltables about Tony Lockett. They were Year (afl season), GM (totals games played in afl season) and GL (total goals kicked in season).  

These are summary statistics i.e. the year summaries (totals) of games played and goals kicked. 

Something you might have noticed with `fitzRoy` is that the `df` is stored as game by game data and not season summaries. 

So how do we create these summary statistics? 

# Step One - Get the dataframe with it all

Data is stored in different places in fitzRoy, df contains the data we are after at the moment

`head(df)` 

```{r}
options(max.print=20)
df<-fitzRoy::get_afltables_stats(start_date = "1897-01-01", end_date = Sys.Date())


```

# Step Two - Select the columns that you want [`select`](https://dplyr.tidyverse.org/reference/select.html)

```{r}
df%>%
  select(First.name, Surname, Season, Goals)

```

# Step Three - Create the grouped data [`group_by`](https://dplyr.tidyverse.org/reference/group_by.html)

Now this might seem a bit weird, because on first look it would seem as though group_by doesn't do anything. 

```{r}

df%>%
  select(First.name, Surname, Season, Goals)%>%
  group_by(Season)
```

It doesn't look like our dataset has changed, but if we look closely we have a new `Groups:   Season [121]` and hopefully what you have noticed is that our dataset `df` has 121 unique values for `Season`. Which we can check below. 

```{r}
length(unique(df$Season))
```

What this does is adds a grouping structure to our data, which means instead of operations acting element wise like they did earlier now operations can happen by group. 

So why would you want to group data? Well lots of interesting things are done on a `group_by` basis. For example we might be interested in goal kicking trends by [`Season`](https://www.foxsports.com.au/afl/afl-goalkicking-continues-to-be-under-scrutiny-as-league-accuracy-falls-to-its-worst-since-1990/news-story/b01d958a5952a86e714e7ef0a41f75d5)
Or 
* by [`round`](https://twitter.com/MatterOfStats/status/997814414832095237) 
* by [`Playing.For`](https://twitter.com/MatterOfStats/status/997814414832095237)
* by [opponent](https://twitter.com/MatterOfStats/status/998017464519675904)
* by [wins and losses](https://twitter.com/MatterOfStats/status/998018492426141696)

We can also `group_by` more than one thing. For example what if we wanted by `Season` and `Round`

```{r}
df%>%
  select(Season,Round, Goals, Behinds)%>%
  group_by(Season, Round)
```
What we can now see is that we have grouped by 2 variables (Season, Round) and have formed 769 groups. `Groups:   Season, Round [2,769]`

# Step Four - Take the grouped data and `summarise` it

So recall earlier we grouped our data by year so lets summarise it. 

```{r}
df%>% 
  select(First.name, Surname, Season, Goals)%>%
  group_by(Season)%>%
  summarise(total_goals=sum(Goals))
```
Ok that doesn't quite seem like what we were thinking we wanted the `total_goals` of Tony Lockett by season, but all we got was Season total goals. 

Well duh! We only did a `group_by` by `Season`, we should have done it by `Season, First.name, Surname` should we have? 

Lets check.

```{r}
df%>%
  select(Season, First.name, Surname, Goals)%>%
  group_by(Season, First.name, Surname)%>%
  summarise(total_goals=sum(Goals))
  
```
This looks a lot better, only issue is that we don't see Tony Lockett but we see all these other blokes like Alb Thomas, he didn't kick 1,360 snags. 

# Step five - `filter` the data

```{r}
df%>%
  select(Season, First.name, Surname, Goals)%>%
  group_by(Season, First.name, Surname)%>%
  summarise(total_goals=sum(Goals))%>%
  filter(First.name=="Tony" , Surname=="Lockett")
```

Now I think I might know what you are thinking. "Weren't we supposed to be plotting average goals by year?"

To do that all we have to do is change `sum` to `mean` and `total_goals` to `average_goals` (these names can be anything but its helpful if they are descriptive).

```{r}
df%>%
  select(Season, First.name, Surname, Goals)%>%
  group_by(Season, First.name, Surname)%>%
  summarise(average_goals=mean(Goals))%>%
  filter(First.name=="Tony" , Surname=="Lockett")
```

# Step Six - `ggplot` our way to a glory

```{r}
df%>%
  select(Season, First.name, Surname, Goals)%>%
  group_by(Season, First.name, Surname)%>%
  summarise(average_goals=mean(Goals))%>%
  filter(First.name=="Tony" , Surname=="Lockett")%>%
  ggplot(aes(x=Season, y=average_goals)) +
  geom_line(color="red") +
  ggtitle("Tony Lockets Average Goals Per Game by Season")

```

# Step 7 - That's cool but I want to compare players

To do this the obvious thing to do, is to just add a player you want to compare to the filter we do this with `%in% c("Player1",("Player2"))` 

```{r}
df%>%
  select(Season, First.name, Surname, Goals)%>%
  group_by(Season, First.name, Surname)%>%
  summarise(average_goals=mean(Goals))%>%
  filter(First.name %in% c("Tony", "Jason") , Surname %in% c("Lockett", "Dunstall"))%>%
  ggplot(aes(x=Season, y=average_goals)) +
  geom_line() +
  ggtitle("Average Goals Per Game by Season")

```

On first glance this doesn't look like its done the right thing we wanted two lines one for Tony Lockett and one for Jason Dunstall. 

Lets look a bit deeper at what we are doing

## Look at the dataframe is it what we wanted?

```{r}
df%>%
  select(Season, First.name, Surname, Goals)%>%
  group_by(Season, First.name, Surname)%>%
  summarise(average_goals=mean(Goals))%>%
  filter(First.name %in% c("Tony", "Jason") , Surname %in% c("Lockett", "Dunstall"))
```

Yes, that looks like the dataframe we want to plot, and looking at the data it seems as though we just have one continous line if we look closely at `Season==1985` we can see that it jumps straight upwards. We know `geom_line()` connects the dots (datapoints) so it looks as though its connecting both values one for Tony Lockett and the other for Jason Dunstall. We could see this a bit more clearly if we looked at a scatter plot.

```{r}
df%>%
    select(Season, First.name, Surname, Goals)%>%
    group_by(Season, First.name, Surname)%>%
    summarise(average_goals=mean(Goals))%>%
    filter(First.name %in% c("Tony", "Jason") , Surname %in% c("Lockett", "Dunstall"))%>%
    ggplot(aes(x=Season, y=average_goals)) +
    geom_point() +
    ggtitle("Average Goals Per Game by Season")
```

## Differentiating the data 

One way to differentiate the data is by colour, i.e. we could just colour Tony Lockett one colour and Jason Dunstall another. 

```{r}
df%>%
    select(Season, First.name, Surname, Goals)%>%
    group_by(Season, First.name, Surname)%>%
    summarise(average_goals=mean(Goals))%>%
    filter(First.name %in% c("Tony", "Jason") , Surname %in% c("Lockett", "Dunstall"))%>%
    ggplot(aes(x=Season, y=average_goals)) +
    geom_point(aes(colour=Surname)) +
    ggtitle("Average Goals Per Game by Season")

```

Then we do our line plot like earlier. 


```{r}
df%>%
    select(Season, First.name, Surname, Goals)%>%
    group_by(Season, First.name, Surname)%>%
    summarise(average_goals=mean(Goals))%>%
    filter(First.name %in% c("Tony", "Jason") , Surname %in% c("Lockett", "Dunstall"))%>%
    ggplot(aes(x=Season, y=average_goals)) +
    geom_line(aes(colour=Surname)) +
    ggtitle("Average Goals Per Game by Season")

```


Another way we could plot them, is have the same graph but instead of both lines on one, we could have them side by side for comparison. This is handy in situations where you want to compare the same thing `average goals per game by season`

This is called [small multiples](https://en.wikipedia.org/wiki/Small_multiple), we do this by using `facet_wrap()`

```{r}
df%>%
    select(Season, First.name, Surname, Goals)%>%
    group_by(Season, First.name, Surname)%>%
    summarise(average_goals=mean(Goals))%>%
    filter(First.name %in% c("Tony", "Jason") , Surname %in% c("Lockett", "Dunstall"))%>%
    ggplot(aes(x=Season, y=average_goals)) +
    geom_line() + facet_wrap(~Surname) + 
    ggtitle("Average Goals Per Game by Season")
```


# Bonus - You did things in a bad way that might lead to problems

Yes I did, my filter would have taken any unique combinations of `First.name` , `Surname` for example if there was a Tony Dunstall or a Jason Lockett they would have come through in the filter. 

We can see this if we changed Dunstall to Smith. 

```{r}

df%>%
    select(Season, First.name, Surname, Goals)%>%
    group_by(Season, First.name, Surname)%>%
    summarise(average_goals=mean(Goals))%>%
    filter(First.name %in% c("Tony", "Jason") , Surname %in% c("Lockett", "Smith"))
    
```

So the best way would be if we had some sort of ID variable. Which luckily in `df` we have an ID column. 

```{r}

df%>%
    select(Season, First.name, Surname,ID, Goals)%>%
    group_by(Season, First.name, Surname, ID)%>%
    summarise(average_goals=mean(Goals))%>%
    filter(First.name %in% c("Tony", "Jason") , Surname %in% c("Lockett", "Dunstall"))
    
```

Now we can see our players have the ID of 990 for Lockett and 632 for Dunstall. 

So instead of filtering by First.name and Surname we can use the ID instead. 

```{r}

df%>%
    select(Season, First.name, Surname,ID, Goals)%>%
    group_by(Season, First.name, Surname, ID)%>%
    summarise(average_goals=mean(Goals))%>%
    filter(ID %in% c(990,632))
```

Then we can now do our same plots as earlier. 
```{r}
df%>%
    select(Season, First.name, Surname,ID, Goals)%>%
    group_by(Season, First.name, Surname, ID)%>%
    summarise(average_goals=mean(Goals))%>%
    filter(ID %in% c(990,632))%>%
    ggplot(aes(x=Season, y=average_goals)) +
    geom_line() + facet_wrap(~Surname) + 
    ggtitle("Average Goals Per Game by Season")

df%>%
    select(Season, First.name, Surname,ID, Goals)%>%
    group_by(Season, First.name, Surname, ID)%>%
    summarise(average_goals=mean(Goals))%>%
    filter(ID %in% c(990,632))%>%
    ggplot(aes(x=Season, y=average_goals)) +
    geom_line(aes(colour=Surname)) +
    ggtitle("Average Goals Per Game by Season")

```