---
title: Couting Pairs
author: Robert Nguyen
date: '2019-04-26'
slug: couting-pairs
categories:
  - fitzRoy
tags:
  - fitzRoy
description: 'Saw a tweet from @matt_pavlich'
---

Saw a [tweet](https://twitter.com/matt_pavlich/status/1121558513111588865) from [@matt_pavlich](https://twitter.com/matt_pavlich) asking twitter roughly how many games he and [David Mundy](https://afltables.com/afl/stats/players/D/David_Mundy.html) have played together. 

Thankfully, you don't have to wonder anymore and you can reproduce the results yourself and do running counts for your favourite players!

```{r}
library(tidyverse)
library(fitzRoy)
df<-fitzRoy::get_afltables_stats(start_date = "1990-01-01", end_date=Sys.Date())
df$matchID<-paste(df$Season, df$Round, df$Home.team, df$Away.team)

df$name<-paste(df$First.name, df$Surname)
df_data<-df%>%filter(Playing.for=="Fremantle")%>%select(name, matchID, Playing.for)
df_data %>%
mutate(n = 1) %>%
spread(name, n, fill=0) %>%
select(-Playing.for, -matchID) %>%
{crossprod(as.matrix(.))} %>%
replace(lower.tri(., diag=T), NA) %>%
reshape2::melt(na.rm=T) %>%
unite('Pair', c('Var1', 'Var2'), sep=", ")%>%
  filter(value>150)%>%
  arrange(desc(value))


```

Some interesting things you might want to do now you have the script and data, you might want to see which pair has played the most together for each team.

Something you might also want to do is look the games that they played together in.

```{r}
df_data%>%
  filter(name %in% c("David Mundy", "Matthew Pavlich")) %>%
  group_by(matchID)%>%
  count(n=n())%>%
  filter(n==2)
```
