---
title: Making a one pager
author: Robert nguyen
date: '2019-01-09'
slug: making-a-one-pager
categories:
  - fitzRoy
  - R
tags:
  - betting
  - DFS
description: 'Maybe handy for fantasy'
---

Why might having a one pager be handy. 

* You might be really into fantasy sports so want to keep information like points given up by position handy.

Lets say you are playing a mate in your favourite fantasy league, having information like points given up by a team to a position tends to be handy. It's nothing new, in fact fantasy sports is big in the US and a site like [rotowire](https://www.rotowire.com/basketball/opp-avg.php) provides this information for fans quickly.

* You might like constructing your own ladders to like a [free kick count](https://twitter.com/triplemfooty/status/1011093422608867329) ladder.

Maybe you like to debate among your mates, maybe you work for a club and want to know where your team sits every week. 


* You might be a punter so having information like [Tony](https://twitter.com/matterofstats?lang=en) presents over in his [matterofstats](http://www.matterofstats.com/mafl-team-dashboards/matterofstats-2018-team-dashboard-for-round-23) might inform your betting strategy. 

No matter the purpose, information is key. But its more than just having the information available. You can construct a lot of the above but it would be very time consuming and probably a bit overwhelming. But that's why [James Day](https://twitter.com/jamesday87) and I have gotten [fitzRoy](https://github.com/jimmyday12/fitzRoy) together so all of these things are relatively quick, easy and can be updated weekly. 

Fantasy sports has grown in popularity recently and people head over to sites like [draftstars](https://www.draftstars.com.au/) and [moneyball](https://moneyball.com.au/landing/generic-fantasy/?utm_source=google&utm_medium=cpc&utm_term=%7Bkeyword%7D&utm_campaign=Google-SEM:Brand&gclid=CjwKCAjwqfDlBRBDEiwAigXUaNRGSQOGENRHLwaFjQwRotNf9T14yS87p8GU8jxgA05RDuReelqi_RoCvGsQAvD_BwE) one of the things people try to do is construct teams for money. 

I'm not really into fantasy sports, but the general gist from chatting to people is this. 

1. You get a salary cap to spend on a team of players
2. There are more expensive and less expensive players, but in general price is correlated positively with points
3. You pick x players and who ever gets the most fantasy points wins!

If we were to look at the US, one of the more popular websites is [rotowire](https://www.rotowire.com/basketball/opp-avg.php) this site people tell me is handy because you are able to see how many points a team gives up to positions. 

If you were doing DFS for AFL one of the questions you might come up with, is "I wonder which team gives up the most fantasy points to forwards".

Well thankfully piecing together a few posts I have done you can do that!

We want to use the fantastic (bit biased here) R package [fitzRoy](https://jimmyday12.github.io/fitzRoy/) which myself and [James Day](https://twitter.com/jamesday87) work on to get the player data from either [afltables](https://afltables.com/afl/afl_index.html) or [footywire](https://www.footywire.com). After this we can use my blog post on [getting player information](https://analysisofafl.netlify.com/data/2018-11-24-getting-player-information/) where we can pull positions of players roughly (forwards, mids, defence and ruck). Then from there we join them together (there will be some missing data but we can fill that in) and lastly we make a similar table to [rotowire](https://www.rotowire.com) which just lists the teams arranged by the average they give up to a certain position. 


```{r}
library(tidyverse)
library(fitzRoy)


df<-fitzRoy::get_footywire_stats(9721:9756)



```

Don't forget to [skim](https://cran.r-project.org/web/packages/skimr/vignettes/Using_skimr.html) your data.

```{r}
skimr::skim(df)

```

Next lets get the players positions and for funsies lets get some other infomation as well.

```{r}
library(rvest)
url<-"https://www.footywire.com/afl/footy/ft_players"


link<-read_html(url)%>%
html_nodes("br+ a , .lnormtop a:nth-child(1)")%>%
  html_attr("href")

url_players<-str_c("https://www.footywire.com/afl/footy/",link)

cbind.fill <- function(...){
    nm <- list(...) 
    nm <- lapply(nm, as.matrix)
    n <- max(sapply(nm, nrow)) 
    do.call(cbind, lapply(nm, function (x) 
        rbind(x, matrix(, n-nrow(x), ncol(x))))) 
}

player_info <- function(x){
  # page <- read_html(x)
  page<-read_html(x)

player<- page%>%
  html_nodes("#playerProfileName")%>%
  html_text()
# player

playing.for<-  page%>%
  html_nodes("#playerProfileTeamDiv a b")%>%
  html_text() %>% as.tibble()
# playing.for
number<-     page%>%
  html_nodes("#playerProfileTeamDiv > b")%>%
  html_text() %>% as.tibble()
# number
weight<-page%>%
  html_nodes("#playerProfileData2")%>%
  html_text()%>%
  str_replace_all("[\r\n]" , "")%>%
  str_squish()%>%
  str_extract(pattern =("(?<=Weight:).*(?=Position:)"))%>%as.tibble()
# weight

height<-page%>%
  html_nodes("#playerProfileData2")%>%
  html_text()%>%
  str_replace_all("[\r\n]" , "")%>%
  str_squish()%>%
  str_extract(pattern =("(?<=Height:).*(?=Weight:)"))%>%as.tibble()
# height
draft_position <- page%>%
  html_nodes("#playerProfileDraftInfo")%>%
  html_text()%>%
  str_replace_all("[\r\n]" , "")%>%
  str_squish()%>%
  str_extract(pattern =("(?<=Drafted: ).*(?=by)"))%>%as.tibble()
# draft_position
club_drafted <- page%>%
  html_nodes("#playerProfileDraftInfo a+ a")%>%
  html_text()%>%str_replace_all("[\r\n]" , "")%>%
  str_squish()%>%
  str_remove(".*by") %>% as.tibble()
# club_drafted
position <-     page%>%
  html_nodes("#playerProfileData2")%>%
  html_text()%>%
  str_replace_all("[\r\n]" , "")%>%
  str_remove(".*Position: ")%>%
  str_squish() %>% as.tibble()

sc_price<-page%>%
  html_nodes("#playerProfileSupercoach")%>%
  html_text()%>%
  str_replace_all("[\r\n]" , "")%>%
  str_squish()%>%
  str_extract(pattern =("(?<=Price:).*(?=Supercoach Profile)"))%>%as.tibble()
# sc_price

af_price<-page%>%
  html_nodes("#playerProfileDreamteam")%>%
  html_text()%>%
  str_replace_all("[\r\n]" , "")%>%
  str_squish()%>%
  str_extract(pattern =("(?<=Price:).*(?=AFL Fantasy Profile)"))%>%as.tibble()
# af_price

player_information <- cbind.fill(player, playing.for, number, weight, height,draft_position, club_drafted, position, sc_price, af_price)

player_information <- as.tibble(player_information)
  
  
  
  
  #combine, name, and make it a tibble
  player_information <- cbind.fill(player, playing.for, number, weight, height,draft_position, club_drafted, position, sc_price, af_price)
  
  player_information <- as.tibble(player_information)

  # print(x)
  # return(x)
  return(player_information)
}
footywire <- purrr::map_df(url_players, player_info)

names(footywire) <- c("player", "club", "number","weight","height",  "draft_position", "club_drafted", "position","sc_price", "af_price")
```

So now we have our dataframes. 

Lets have a looksie

```{r}
head(footywire)
head(df)

```

What we can see here is that if we were to attempt to join on team and player name that the team names do not align between the freshly scraped data and the dataset within [fitzRoy](https://github.com/jimmyday12/fitzRoy) so lets replace the team names in the freshly scraped footywire data to match those existing already within [#fitzRoy](https://github.com/jimmyday12/fitzRoy).

Too see how its an issue we can use `unique` 

```{r}
unique(footywire$club)
unique(df$Team)

```

```{r}
footywire<-footywire%>%
  mutate(club=replace(club, club=="Richmond Tigers","Richmond" ))%>%
  mutate(club=replace(club, club=="Geelong Cats" ,"Geelong"  ))%>%
  mutate(club=replace(club, club=="St Kilda Saints" ,"St Kilda"  ))%>%
  mutate(club=replace(club, club=="Brisbane Lions" ,"Brisbane"))%>%
  mutate(club=replace(club, club=="Collingwood Magpies" ,"Collingwood" ))%>%
  mutate(club=replace(club, club=="West Coast Eagles" ,"West Coast" ))%>%
  mutate(club=replace(club, club=="Gold Coast Suns" ,"Gold Coast" ))%>%
  mutate(club=replace(club, club=="North Melbourne Kangaroos","North Melbourne"  ))%>%
  mutate(club=replace(club, club=="Sydney Swans" , "Sydney"  ))%>%
  mutate(club=replace(club, club=="Essendon Bombers" ,"Essendon" ))%>%
  mutate(club=replace(club, club=="Port Adelaide Power","Port Adelaide" ))%>%
  mutate(club=replace(club, club=="Adelaide Crows" ,"Adelaide" ))%>%
  mutate(club=replace(club, club=="Melbourne Demons" ,"Melbourne"  ))%>%
  mutate(club=replace(club, club=="Fremantle Dockers" ,"Fremantle"  ))%>%
  mutate(club=replace(club, club=="Hawthorn Hawks", "Hawthorn"  ))%>%
  mutate(club=replace(club, club=="GWS Giants"  ,"GWS" ))%>%
  mutate(club=replace(club, club== "Footscray" ,"Western Bulldogs" ))%>%
  mutate(club=replace(club, club=="Carlton Blues"    ,"Carlton"  ))

```

Then we check it. I find using `str_sort` helps to see the output on top of each other. 

```{r}
str_sort(unique(footywire$club))
str_sort(unique(df$Team))

```

From the footywire data, we want to take the player names, club and position and then join this onto the `player_stats` dataset from [fitzRoy](https://github.com/jimmyday12/fitzRoy). 

```{r}
footywire_positions<-select(footywire, player, club, position)
footywire_2019<-df

dataset<-left_join(footywire_2019, footywire_positions, by=c("Player"="player", "Team"="club"))
```

Issue here is that the hyphen names don't match so we got to change those aswell. 

We can view these problem rows as follows.

```{r}
dataset[rowSums(is.na(dataset)) > 0,]

```

So what is the issue? 

If we go to say [Anthony McDonald-Tipungwuti](https://www.footywire.com/afl/footy/pg-essendon-bombers--anthony-mcdonald-tipungwuti) page where we take his player information we can see that his name is listed as Anthony McDonald-Tipungwuti, however if we were to go to an [Essendon Bombers](https://www.footywire.com/afl/footy/th-essendon-bombers) game, such as their one vs the [Giants](https://www.footywire.com/afl/footy/ft_match_statistics?mid=9728) we can see that his name is Anthony M-Tipungwuti which is what is in the [fitzRoy](https://github.com/jimmyday12/fitzRoy) dataset.


So lets change the names in the scrape to match the names within [fitzRoy](https://github.com/jimmyday12/fitzRoy).

```{r}
footywire<-footywire%>%
  mutate(player=replace(player, player=="Sam Petrevski-Seton","Sam P-Seton" ))%>%
   mutate(player=replace(player, player=="Alex Neal-Bullen","Alex N-Bullen" ))%>%
   mutate(player=replace(player, player=="Sam Powell-Pepper","Sam P-Pepper" ))%>%
   mutate(player=replace(player, player=="Darcy Byrne-Jones","Darcy B-Jones" ))%>%
   mutate(player=replace(player, player=="George Horlin-Smith","George H-Smith" ))%>%
    mutate(player=replace(player, player=="Anthony McDonald-Tipungwuti","Anthony M-Tipungwuti" ))%>%
   mutate(player=replace(player, player=="Luke Davies-Uniacke","Luke D-Uniacke" ))%>%
  mutate(player=replace(player, player=="Cameron Ellis-Yolmen","Cameron E-Yolmen" )) %>%
  mutate(player=replace(player, player=="Ed Vickers-Willis","Ed V-Willis" ))
  
footywire_positions<-select(footywire, player, club, position)  
  
dataset<-left_join(footywire_2019, footywire_positions, by=c("Player"="player", "Team"="club"))

```
Now lets do some plotting!

Suppose we wanted to see which players did well playing the West Coast Eagles, we could do a simple plot of position by their fantasy scores. 

```{r}
dataset%>%
    filter(Opposition=="West Coast")%>%
    ggplot(aes(y=AF, x=as.factor(position)))+
    geom_point()+ 
    geom_text(aes(label=Player),check_overlap = TRUE, size=2) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

```{r}
dataset%>%
    ggplot(aes(y=AF, x=as.factor(position)))+
    geom_violin()+ 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+ facet_wrap( ~ Opposition, ncol=6)

```



