---
title : "Bradley Terry"
description : "Building a quick Bradley Terry model for AFL"
author : "Robert Nguyen"
date : 2018-06-22
tags : ["regression", "fitzRoy","modelling"]

---


You might be reading other sites and see lots of posts on [ELO](https://thearcfooty.com/2016/12/29/introducing-the-arcs-ratings-system/) [models](http://figuringfooty.com/2015/10/09/the-simpelo-team-ratings/) this post isn't about building an ELO model yourself that's another post. This post is about how to apply [Bradley Terry](https://en.wikipedia.org/wiki/Bradley%E2%80%93Terry_model) models for AFL

# Why Bother?

Each of the 18 footy teams plays 22 games a year, we can see who has played who and who won each game. We could just use the [AFL ladder](http://www.afl.com.au/ladder) but each team doesn't play each other equally some teams have have easier [schedules](http://www.matterofstats.com/mafl-stats-journal/2016/10/27/an-analysis-of-strength-of-schedule-for-the-2018-fixture) than others. What strength of shedule means is that we can't use the ladder as the best measure of teams as some times might have a worse win/loss record but simply have had a tougher draw than opposing teams who might have had it a lot easier. 

A model based approach can help address this problem. 

# Introducing Bradley Terry

A simple Bradley Terry model treats outcomes of games as an independent Bernoulli random variable with a Bernoulli distribution $p_{ij}$ 

The log odds corresponding to the probability $p_{ij}$  that team i beats team j is

$log\frac{p_{ij}}{1-p_{ij}} = \beta{i} - \beta{j}$

The problem is that this is over-parametized which means that its exactly the same if we were to add a fixed constant to all the values of $\beta{i}$

# What about HGA

The model as it stands now doesn't have home ground advantage. We can incorporate that by including an intercept term $\alpha$ 

$log\frac{p_{ij}}{1-p_{ij}} = \alpha + \beta{i} - \beta{j}$

By rearranging the equation we can see how it increases the log-odds of the home team winning by a constant $\alpha$ 


```{r}
library(fitzRoy)
library(tidyverse)

df<-fitzRoy::get_match_results()
num_teams=18
df1<-df%>%filter(Season==2017)%>%
  mutate(Y=if_else(Home.Points>Away.Points, 1, 0))

teams=unique(df1$Home.Team)
df1$Home.Team[df1$Home.Team == "Richmond"] <- 1
df1$Home.Team[df1$Home.Team == "Essendon"] <- 2
df1$Home.Team[df1$Home.Team == "Port Adelaide"] <- 3
df1$Home.Team[df1$Home.Team == "Hawthorn"] <- 4
df1$Home.Team[df1$Home.Team ==  "Gold Coast"] <- 5
df1$Home.Team[df1$Home.Team == "GWS"] <- 6
df1$Home.Team[df1$Home.Team == "Melbourne"] <- 7
df1$Home.Team[df1$Home.Team ==  "West Coast"] <- 8
df1$Home.Team[df1$Home.Team == "Adelaide"] <- 9
df1$Home.Team[df1$Home.Team == "North Melbourne"] <- 10
df1$Home.Team[df1$Home.Team == "Carlton"] <- 11
df1$Home.Team[df1$Home.Team == "Collingwood"] <- 12
df1$Home.Team[df1$Home.Team == "Brisbane Lions"] <- 13
df1$Home.Team[df1$Home.Team == "Fremantle"] <- 14
df1$Home.Team[df1$Home.Team == "Footscray"] <- 15
df1$Home.Team[df1$Home.Team == "Sydney"] <- 16
df1$Home.Team[df1$Home.Team == "Geelong"] <- 17
df1$Home.Team[df1$Home.Team == "St Kilda"] <- 18

df1$Home.Team<-as.integer(df1$Home.Team)


df1$Away.Team[df1$Away.Team == "Richmond"] <- 1
df1$Away.Team[df1$Away.Team == "Essendon"] <- 2
df1$Away.Team[df1$Away.Team == "Port Adelaide"] <- 3
df1$Away.Team[df1$Away.Team == "Hawthorn"] <- 4
df1$Away.Team[df1$Away.Team ==  "Gold Coast"] <- 5
df1$Away.Team[df1$Away.Team == "GWS"] <- 6
df1$Away.Team[df1$Away.Team == "Melbourne"] <- 7
df1$Away.Team[df1$Away.Team ==  "West Coast"] <- 8
df1$Away.Team[df1$Away.Team == "Adelaide"] <- 9
df1$Away.Team[df1$Away.Team == "North Melbourne"] <- 10
df1$Away.Team[df1$Away.Team == "Carlton"] <- 11
df1$Away.Team[df1$Away.Team == "Collingwood"] <- 12
df1$Away.Team[df1$Away.Team == "Brisbane Lions"] <- 13
df1$Away.Team[df1$Away.Team == "Fremantle"] <- 14
df1$Away.Team[df1$Away.Team == "Footscray"] <- 15
df1$Away.Team[df1$Away.Team == "Sydney"] <- 16
df1$Away.Team[df1$Away.Team == "Geelong"] <- 17
df1$Away.Team[df1$Away.Team == "St Kilda"] <- 18

df1$Away.Team<-as.integer(df1$Away.Team)

loglik = function(theta, Home.Team, Away.Team, Y) {
  alpha = theta[1]
  beta = c(0, theta[-1])
  params = alpha + beta[Home.Team] - beta[Away.Team]
  return(sum(Y * params - log(1 + exp(params))))
}


theta0 = rep(0, num_teams)

result = optim(theta0, loglik,
               Home=df1$Home.Team, Away=df1$Away.Team, Y=df1$Y,
               method='BFGS', control=list('fnscale'=-1))

coefs = c(0, result$par[-1])
ranking = order(coefs, decreasing=TRUE)
ranking



```

# So what is our HGA using a Bradley Terry Model?

```{r}
result$par[1]
exp(-result$par[1])

```


# Is HGA significant? 

```{r}

loglik_noalpha = function(theta, Home, Away, Y) {
beta = c(0, theta)
params = beta[Home] - beta[Away]
return(sum(Y * params - log(1 + exp(params))))
}
theta0 = rep(0, num_teams - 1)
result_noalpha = optim(theta0, loglik_noalpha,
Home=df1$Home.Team, Away=df1$Away.Team, Y=df1$Y,
method='BFGS', control=list('fnscale'=-1))

print(result$value)

print(result_noalpha$value)

statistic = -2 * (result_noalpha$value - result$value)
p_value = 1 - pchisq(statistic, df=1)
print(statistic)

print(p_value)

```

Some problems with Bradley-Terry here:

*  I haven't added in changes in strength, this means equal weight to recent and past games
* Using logistic while makes the calculations quick and easy, it has an important implication

$ P(i beats j)=2/3, P(j beats k)= 2/3 then P(i beats k)=4/5$ 
* To use this for prediction for a season, would need to recompute MLEs after each game



